\singlespacing \vspace{-1.6cm} \LARGE Covariation among reproductive traits in flowering plants determine interactions with floral visitors

### 2022-05-30 Manuscript/Lanuza_et_al_2021.Rmd
Some edits
edits

We explored the association between the different quantitative plant traits with a phylogenetically informed Principal Component Analysis (pPCA). We did not include the quantitative variables of flower length and inflorescence width because they were highly and moderately correlated to flower width respectively (Pearson's correlation = 0.72, *P* \< 0.01 and Pearson's correlation = 0.36, *P* \< 0.01), and thus we avoided overemphasizing flower size on the spectrum of trait variation. Although qualitative traits were not included in the dimensionality reduction analysis, we also investigated the statistical association of the different qualitative traits with the main axes of trait variation with the help of an Anova and a Tukey test. For this, we first implemented a linear model (one for each principal component) where they were the response variable and the different qualitative variables the predictors. Prior to the analyses, we excluded outliers and standardized the data. Due to the high sensitivity of dimensionality reduction to outliers [@serneels2008; @legendre2012], we excluded values within the 2.5th--97.5th percentile range, and thus our final dataset had 1,236 species. Then, we log transformed the variables to reduce the influence of outliers and z-transformed (X= 0, SD=1) so that all variables were within the same numerical range. We performed the pPCA using the function *phyl.pca* from the package *phytools* [@revell2012] with the method lambda ($\lambda$) that calculates the phylogenetic correlation between 0 (phylogenetic independence) and 1 (shared evolutionary history) and we implemented the mode covariance because values for each variables were on the same scale following transformation [@abdi2010]. Moreover, to corroborate that our imputation of missing values did not affect our results, we conducted a pPCA on the full dataset without missing values (Fig. S3). We found little difference between the explained variance with the imputed dataset (51.08%) and the dataset without missing values (52.87%). In addition, the loadings on each principal component had a similar contribution and correlation patterns, with the exception of plant height which showed slight variations between the imputed and non-imputed dataset. Finally, we conducted an additional phylogenetic informed principal component analysis for the subset of species with pollen and nectar quantity. For this, we included all quantitative traits considered in the main pPCA plus pollen grains and microlitres of nectar per flower.

@@

The phylogenetically informed principal component analysis captured by the first two and three axes 51.8% and 70.97% of the reproductive trait variation, respectively (Fig. 1 and Fig. S6). The first principal component (PC1) represented 26.72% of the trait variation and indicated a negative correlation between flower number and flower size. The main contributing traits to PC1 were plant height, flower number, ovule number and flower size (loadings \> \|0.5\|; Table S3) but style length also contributed moderately to PC1 (loading = -0.33). One end of this axis comprised species with high investment in flower number and plant height but small flower size, short style length and low ovule number. For instance, on this end of the spectrum we find the species *Cornus florida* which has approximately a total of 10.000 flowers, an average height of 7.5m, flowers of 3mm wide, a style length of 3.5mm and a total of 2 ovules per flower. The other end of this spectrum comprised species that were short in height and invested in large flowers, long styles, many ovules, but few flowers. For instance, on this side of the spectrum we find the species *Petunia axillaris* which has approximately 10 flowers per plant, a height of 0.5m, flowers over 50mm wide, styles of 25mm and over 200 ovules per flower. The second principal component (PC2) represented 25.05% of the trait variation and indicated the variation from low to high autonomous selfing, or in other words, high to low pollinator dependence, respectively. The main driver of trait variation on PC2 was autonomous selfing (loading = 0.85) but the other traits (except ovule number) also made moderate contributions (loadings from 0.27 to 0.4; Table S3).

We found that high pollinator dependence was associated with larger and a higher number of flowers, greater plant height and longer styles. In contrast, species with high levels of autonomous selfing tended to have fewer and smaller flowers, had shorter styles and were shorter in height. Further, PC3 explained a considerable amount of trait variability (19.17%) and the main contributors to this axis were style length (loading = -0.66) and the degree of autonomous selfing (loading = -0.51). The remaining traits, apart from ovule number, were moderately correlated to changes on PC3 (loadings from -0.23 to -0.46; Table S3). Thus, because style length was correlated with all traits on PC3 and was the main driver of trait variation, we refer to this axis as the 'style length trade-off'. Further, the pPCA with the subset of species that had nectar and pollen quantity data showed that nectar quantity (microlitres of nectar per flower) and pollen grains per flower were positively associated with flower size, style length and ovule number but negatively associated with flower number (PC1, 26.82%; Fig. S7). This pPCA explained similar variance with the first two principal components (45.52%) and similar associations of traits despite some variability in the loadings (Table S4).

@@

We found a strong phylogenetic signal (*P* \< 0.01) in all quantitative traits (Table S5). The traits that showed the highest phylogenetic signal were ovule number ($\lambda$ = 1), pollen grains per flower ($\lambda$ = 1) and plant height ($\lambda$ = 0.96), followed by flower length ($\lambda$ = 0.75), flower width ($\lambda$ = 0.73), number of flowers per plant ($\lambda$ = 0.69) and nectar concentration ($\lambda$ = 0.65). The traits that showed a moderate phylogenetic signal were inflorescence width ($\lambda$ = 0.57), style length ($\lambda$ = 0.49) and autonomous selfing ($\lambda$ = 0.34). Finally, microliters of nectar per flower showed the lowest phylogenetic signal of all traits ($\lambda$ = 0.14).

@@

  plot <- plot + xlab(paste("Flower number - flower size trade-off ", "(",(percentage[1]),"%",")", sep = "")) #XLAB
  plot <- plot + ylab(paste("Pollinator dependence trade-off ", "(",(percentage[2]),"%",")", sep = "")) #YLAB

@@

```{=tex}
\vspace{5mm}
\blandscape
```
```{r, echo=FALSE, message=FALSE, cache=FALSE, results='hide', warning=FALSE, fig.height=14, fig.width=27,fig.cap= "**Figure 3** Fitted posterior estimates of the presence-absence of interaction (a, b and c) and number of visits (d, e and f) of the different floral visitor guilds in relation to the main axes of trait variation (PC1, PC2 and PC3). PC1 represents the flower number - flower size trade-off, PC2 represents the pollinator dependence trade-off and PC3, the style length trade-off. For visualization purposes, due to large differences between number of visits of bees and the rest of guilds, the number of visits was log-transformed (Y-axis of lower panel)."}

>>>>

We explored the association between the different quantitative plant traits with a phylogenetically informed Principal Component Analysis (pPCA). We did not include the quantitative variables of flower length and inflorescence width because they were highly and moderately correlated to flower width respectively (Pearson's correlation = 0.72, *P* \< 0.01 and Pearson's correlation = 0.36, *P* \< 0.01), and thus we avoided overemphasizing flower size on the spectrum of trait variation. Prior to the analyses, we excluded outliers and standardized the data. Due to the high sensitivity of dimensionality reduction to outliers [@serneels2008; @legendre2012], we excluded values outside the 2.5th--97.5th percentile range, and thus our final dataset had 1,236 species. Then, we log transformed the variables to reduce the influence of outliers and z-transformed (X= 0, SD=1) so that all variables were within the same numerical range as indicated for principal component analysis [@legendre2012]. Although qualitative traits were not included in the dimensionality reduction analysis, we also investigated the statistical association of the different qualitative traits with the main axes of trait variation with the help of an Anova and a Tukey test. We performed the pPCA using the function *phyl.pca* from the package *phytools* [@revell2012] with the method lambda ($\lambda$) that calculates the phylogenetic correlation between 0 (phylogenetic independence) and 1 (shared evolutionary history) and we implemented the mode covariance because values for each variables were on the same scale following transformation [@abdi2010]. Moreover, to corroborate that our imputation of missing values did not affect our results, we conducted a pPCA on the full dataset without missing values (Fig. S3). We found little difference between the explained variance with the imputed dataset (51.08%) and the dataset without missing values (52.87%). In addition, the loadings on each principal component had a similar contribution and correlation patterns, with the exception of plant height which showed slight variations between the imputed and non-imputed dataset. Finally, we conducted an additional phylogenetic informed principal component analysis for the subset of species with pollen and nectar quantity. For this, we included all quantitative traits considered in the main pPCA plus pollen grains and microlitres of nectar per flower.

@@

The phylogenetically informed principal component analysis captured by the first two and three axes 51.8% and 70.97% of the reproductive trait variation, respectively (Fig. 1 and Fig. S6). The first principal component (PC1) represented 26.72% of the trait variation and indicated a negative correlation between flower number and flower size ('flower number - flower size'). The main contributing traits to PC1 were plant height, flower number, ovule number and flower size (loadings \> \|0.5\|; Table S3) but style length also contributed moderately to PC1 (loading = -0.33). One end of this axis comprised species with high investment in flower number and plant height but small flower size, short style length and low ovule number. For instance, on this end of the spectrum we find the species *Cornus florida* which has approximately a total of 10.000 flowers, an average height of 7.5m, flowers of 3mm wide, a style length of 3.5mm and a total of 2 ovules per flower. The other end of this spectrum comprised species that were short in height and invested in large flowers, long styles, many ovules, but few flowers. For instance, on this side of the axis we find the species *Petunia axillaris* that has approximately 10 flowers per plant, a height of 0.5m, flowers over 50mm wide, styles of 25mm and over 200 ovules per flower. The second principal component (PC2) represented 25.05% of the trait variation and indicated the variation from low to high autonomous selfing, or in other words, high to low pollinator dependence, respectively. The main driver of trait variation on PC2 was autonomous selfing (loading = 0.85) but the other traits (except ovule number) also made moderate contributions (loadings from 0.27 to 0.4; Table S3). In general terms, species with high pollinator dependence were associated with larger and a higher number of flowers, greater plant height and longer styles. A species example of this side of the spectrum is *Zuccagnia punctata* that is a self incompatible shrub which depends totally on floral visitors, it has approximately 1500 flowers per plant, 3m height and a style length of 20mm. In contrast, species with low pollinator dependence tended to have fewer and smaller flowers, shorter plant height and shorter styles. For instance, on this side of the spectrum we find the species *Veronica peregrina* which is a self compatible herb that is thought to be almost a complete selfer, it has around 20 flowers per plant, a height of 0.2m and a style length of 0.25mm. Further, PC3 explained a considerable amount of trait variability (19.17%) and the main contributors to this axis were style length (loading = -0.66) and the degree of autonomous selfing (loading = -0.51). The remaining traits, apart from ovule number, were moderately correlated to changes on PC3 (loadings from -0.23 to -0.46; Table S3). In addition, the pPCA with the subset of species that had nectar and pollen quantity data showed that nectar quantity (microlitres of nectar per flower) and pollen grains per flower were positively associated with flower size, style length and ovule number but negatively associated with flower number (PC1, 26.82%; Fig. S7). This pPCA explained similar variance with the first two principal components (45.52%) and similar associations of traits despite some variability in the loadings (Table S4).

@@

We found a strong phylogenetic signal (*P* \< 0.01) for most quantitative traits (Table S5). The traits that showed the highest phylogenetic signal were ovule number ($\lambda$ = 1), pollen grains per flower ($\lambda$ = 1) and plant height ($\lambda$ = 0.96), followed by flower length ($\lambda$ = 0.75), flower width ($\lambda$ = 0.73), number of flowers per plant ($\lambda$ = 0.69) and nectar concentration ($\lambda$ = 0.65). The traits that showed a moderate phylogenetic signal were inflorescence width ($\lambda$ = 0.57), style length ($\lambda$ = 0.49) and autonomous selfing ($\lambda$ = 0.34). Finally, microliters of nectar per flower showed the lowest phylogenetic signal of all traits ($\lambda$ = 0.14).

@@

  plot <- plot + xlab(paste("Flower number - flower size axis ", "(",(percentage[1]),"%",")", sep = "")) #XLAB
  plot <- plot + ylab(paste("Pollinator dependence axis ", "(",(percentage[2]),"%",")", sep = "")) #YLAB

@@

```{=tex}
\vspace{5mm}
\blandscape
```
```{r, echo=FALSE, message=FALSE, cache=FALSE, results='hide', warning=FALSE, fig.height=14, fig.width=27,fig.cap= "**Figure 3** Fitted posterior estimates of the presence-absence of interaction (a, b and c) and number of visits (d, e and f) of the different floral visitor guilds in relation to the main axes of trait variation (PC1, PC2 and PC3). PC1 represents the flower number - flower size axis, PC2 represents the pollinator dependence axis and PC3, the style length - pollinator dependence axis. For visualization purposes, due to large differences between number of visits of bees and the rest of guilds, the number of visits was log-transformed (Y-axis of lower panel)."}

### 2022-05-30 Manuscript/Lanuza_et_al_2021.tex
edits

We selected a total of 19 different functional traits that comprised both reproductive and life history traits (see Table 1). From these, 16 were reproductive traits that consisted of 13 floral traits and 3 reproductive biology traits. Floral traits included morphological traits (e.g., style length) but also floral reward (e.g., pollen quantity) and floral display (e.g., number of flowers) related traits. Reproductive biology traits indicated the reproductive system of the plant and included breeding, mating and compatibility system. The 3 remaining traits were life history traits that included plant height, lifespan and life form and are commonly used to characterize the fast-slow continuum of plant trait variation (e.g., short-lived versus perennial species). For each plant species, we undertook an extensive literature and online search for all traits across a wide range of resources (plant databases, online floras, books, journals and images). From a total of 30,120 possible cells considering all traits and plant species (20 columns × 1,506 species), we were able to fill 24,341 cells (80.8\% of the dataset, see Fig. S2 for missing values information for each trait). An extended description of each trait and how it was obtained can be found in Appendix S1.

@@

We explored the trade-offs between the different quantitative plant traits with a phylogenetically informed Principal Component Analysis (pPCA). We did not include the quantitative variables of flower length and inflorescence width because they were highly and moderately correlated to flower width respectively (Pearson's correlation = 0.72, \emph{P} \textless{} 0.01 and Pearson's correlation = 0.36, \emph{P} \textless{} 0.01), and thus we avoided overemphasizing flower size on the spectrum of trait variation. Although qualitative traits were not included in the dimensionality reduction analysis, we also investigated the association of the different qualitative traits with the main axes of trait variation. Prior to the analyses, we excluded outliers and standardized the data. Due to the high sensitivity of dimensionality reduction to outliers, we excluded values within the 2.5th--97.5th percentile range (\protect\hyperlink{ref-legendre2012}{Legendre \& Legendre, 2012}), and thus our final dataset had 1,236 species. Then, we log transformed the variables to reduce the influence of outliers and z-transformed (X= 0, SD=1) so that all variables were within the same numerical range. We performed the pPCA using the function \emph{phyl.pca} from the package \emph{phytools} (\protect\hyperlink{ref-revell2012}{Revell, 2012}) with the method lambda (\(\lambda\)) that calculates the phylogenetic correlation between 0 (phylogenetic independence) and 1 (shared evolutionary history) and we implemented the mode covariance because values for each variables were on the same scale following transformation (\protect\hyperlink{ref-abdi2010}{Abdi \& Williams, 2010}). Moreover, to corroborate that our imputation of missing values did not affect our results, we conducted a pPCA on the full dataset without missing values (Fig. S3). We found little difference between the explained variance with the imputed dataset (51.08\%) and the dataset without missing values (52.87\%). In addition, the loadings on each principal component had a similar contribution and correlation patterns, with the exception of plant height which showed slight variations between the imputed and non-imputed dataset. Finally, we conducted an additional phylogenetic informed principal component analysis for the subset of species with pollen and nectar quantity. For this, we included all quantitative traits considered in the main pPCA plus pollen grains and microlitres of nectar per flower.

@@

We first investigated how the different groups of floral visitors interacted along the main axes of trait variation (see below `visitation patterns' section). This was done for the binary version of the networks that assumes equal weight across interactions and for the weighted version with the number of visits of floral visitors to individual flowers that account for the intensity of the interaction. Although floral visitors are not always pollinators and number of visits does not consider each pollinator species efficiency (\protect\hyperlink{ref-ballantyne2015}{Ballantyne et al., 2015}), the number of visits can provide valuable information of the contribution of floral visitors to pollination (\protect\hyperlink{ref-vazquez2005}{Vázquez et al., 2005}, \protect\hyperlink{ref-vazquez2012}{2012}). Second, we investigated how the main axes of trait variation and individual traits influence plant species' roles within networks using a set of complementary interaction network metrics: number of visits, normalized degree and specialization (see below `plant species network roles' section).

@@

We investigated whether different quantitative traits determined plant species network roles using Bayesian modelling and regression trees. For this, we selected simple and complementary species-level network metrics commonly applied in bipartite network studies (\protect\hyperlink{ref-dormann2008}{Dormann et al., 2008}) with a straightforward ecological interpretation relevant to our research goals. The different plant species-level metrics were: (i) sum of visits per plant species; (ii) normalized degree, calculated as the number of links per plant species divided by the total possible number of partners; and (iii) specialization (d') (\protect\hyperlink{ref-bluthgen2006}{Blüthgen et al., 2006}), which measures the deviation of an expected random choice of the available interaction partners and ranges between 0 (maximum generalization) and 1 (maximum specialization). Normalized degree and specialization were calculated with the \emph{specieslevel} function from the R package \emph{bipartite} (\protect\hyperlink{ref-dormann2008}{Dormann et al., 2008}).

@@

The phylogenetically informed principal component analysis captured by the first two and three axes 51.8\% and 70.97\% of trait variation, respectively (Fig. 1 and Fig. S6) and had a phylogenetic correlation (\(\lambda\)) of 0.76. The first principal component (PC1) represented 26.72\% of the trait variation and indicated a trade-off between flower number and flower size. We refer to this axis as the `flower number - flower size trade-off,' as already described in previous studies (\protect\hyperlink{ref-kettle2011}{Kettle et al., 2011}; \protect\hyperlink{ref-sargent2007}{Sargent et al., 2007}). Hence, one end of the spectrum comprised species with high investment in flower number and plant height but small flower size, short style length and low ovule number. The other end of this spectrum comprised species that were short in height and invested in large flowers, long styles, many ovules, but few flowers. The main contributing traits to PC1 were plant height, flower number, ovule number and flower size (loadings \textgreater{} \textbar0.5\textbar; Table S3) but style length also contributed moderately to PC1 (loading = -0.33). The second principal component (PC2) represented 25.05\% of the trait variation and indicated a trade-off between low and high pollinator dependence. We refer to this axis as the `pollinator dependence trade-off.' The main driver of trait variation on PC2 was autonomous selfing (loading = 0.85) but the other traits (except ovule number) also made moderate contributions (loadings from 0.27 to 0.4; Table S3). We found that high pollinator dependence was associated with larger and a higher number of flowers, greater plant height and longer styles. In contrast, species with high levels of autonomous selfing tended to have fewer and smaller flowers, had shorter styles and were shorter in height. Further, PC3 explained a considerable amount of trait variability (19.17\%) and the main contributors to this axis were style length (loading = -0.66) and the degree of autonomous selfing (loading = -0.51). The remaining traits, apart from ovule number, were moderately correlated to changes on PC3 (loadings from -0.23 to -0.46; Table S3). Thus, because style length was correlated with all traits on PC3 and was the main driver of trait variation, we refer to this axis as the `style length trade-off.' Further, the pPCA with the subset of species that had nectar and pollen quantity data showed that nectar quantity (microlitres of nectar per flower) and pollen grains per flower were positively associated with flower size, style length and ovule number but negatively associated with flower number (PC1, 26.82\%; Fig. S7). This pPCA explained similar variance with the first two principal components (45.52\%) and similar associations of traits despite some variability in the loadings (Table S4).

@@

We found a strong phylogenetic signal (\emph{P} \textless{} 0.01) in all quantitative traits (Table S5). The traits that showed the highest phylogenetic signal were ovule number (\(\lambda\) = 1), pollen grains per flower (\(\lambda\) = 1) and plant height (\(\lambda\) = 0.96), followed by flower length (\(\lambda\) = 0.75), flower width (\(\lambda\) = 0.73), number of flowers per plant (\(\lambda\) = 0.69) and nectar concentration (\(\lambda\) = 0.65). The traits that showed a moderate phylogenetic signal were inflorescence width (\(\lambda\) = 0.57), style length (\(\lambda\) = 0.49) and autonomous selfing (\(\lambda\) = 0.34). Finally, microliters of nectar per flower showed the lowest phylogenetic signal of all traits (\(\lambda\) = 0.14).

@@

@@

\begin{figure}
\centering
\includegraphics{output/figures/unnamed-chunk-4-1.pdf}
\caption{\label{fig:unnamed-chunk-4}\textbf{Figure 3} Fitted posterior estimates of the presence-absence of interaction (a, b and c) and number of visits (d, e and f) of the different floral visitor guilds in relation to the main axes of trait variation (PC1, PC2 and PC3). PC1 represents the flower number - flower size trade-off, PC2 represents the pollinator dependence trade-off and PC3, the style length trade-off. For visualization purposes, due to large differences between number of visits of bees and the rest of guilds, the number of visits was log-transformed (Y-axis of lower panel).}
\end{figure}

>>>>

We selected a total of 19 different functional traits that comprised both reproductive and life history traits (see Table 1). From these, 16 were reproductive traits that consisted of 13 floral traits and 3 reproductive biology traits. Floral traits included morphological traits (e.g., style length) but also floral reward (e.g., pollen quantity) and floral display (e.g., number of flowers) related traits. Reproductive biology traits indicated the reproductive system of the plant and included breeding, mating and compatibility system. The 3 remaining traits were life history traits that are commonly used to characterize the fast-slow continuum of plant trait variation (i.e., plant height, lifespan and life form). For each plant species, we undertook an extensive literature and online search for all traits across a wide range of resources (plant databases, online floras, books, journals and images). From a total of 30,120 possible cells considering all traits and plant species (20 columns × 1,506 species), we were able to fill 24,341 cells (80.8\% of the dataset, see Fig. S2 for missing values information for each trait). An extended description of each trait and how it was obtained can be found in Appendix S1.

@@

We explored the association between the different quantitative plant traits with a phylogenetically informed Principal Component Analysis (pPCA). We did not include the quantitative variables of flower length and inflorescence width because they were highly and moderately correlated to flower width respectively (Pearson's correlation = 0.72, \emph{P} \textless{} 0.01 and Pearson's correlation = 0.36, \emph{P} \textless{} 0.01), and thus we avoided overemphasizing flower size on the spectrum of trait variation. Although qualitative traits were not included in the dimensionality reduction analysis, we also investigated the statistical association of the different qualitative traits with the main axes of trait variation. For this, we first implemented a linear model (one for each principal component) where the principal components were the response variable and the different qualitative variables the predictors. Finally, we explore the statistical association with an Anova and the differences between categories for each categorical variable with a Tukey test. Prior to the analyses, we excluded outliers and standardized the data. Due to the high sensitivity of dimensionality reduction to outliers (\protect\hyperlink{ref-legendre2012}{Legendre \& Legendre, 2012}; \protect\hyperlink{ref-serneels2008}{Serneels \& Verdonck, 2008}), we excluded values within the 2.5th--97.5th percentile range, and thus our final dataset had 1,236 species. Then, we log transformed the variables to reduce the influence of outliers and z-transformed (X= 0, SD=1) so that all variables were within the same numerical range. We performed the pPCA using the function \emph{phyl.pca} from the package \emph{phytools} (\protect\hyperlink{ref-revell2012}{Revell, 2012}) with the method lambda (\(\lambda\)) that calculates the phylogenetic correlation between 0 (phylogenetic independence) and 1 (shared evolutionary history) and we implemented the mode covariance because values for each variables were on the same scale following transformation (\protect\hyperlink{ref-abdi2010}{Abdi \& Williams, 2010}). Moreover, to corroborate that our imputation of missing values did not affect our results, we conducted a pPCA on the full dataset without missing values (Fig. S3). We found little difference between the explained variance with the imputed dataset (51.08\%) and the dataset without missing values (52.87\%). In addition, the loadings on each principal component had a similar contribution and correlation patterns, with the exception of plant height which showed slight variations between the imputed and non-imputed dataset. Finally, we conducted an additional phylogenetic informed principal component analysis for the subset of species with pollen and nectar quantity. For this, we included all quantitative traits considered in the main pPCA plus pollen grains and microlitres of nectar per flower.

@@

First, we investigated how the different groups of floral visitors interacted along the main axes of reproductive trait variation (see below `visitation patterns' section) with the help of qualitative and quantitative information of plant-pollinator interactions. For this, we used as qualitative information the binary version of the networks (presence-absence of interaction) that assumes equal weight across interactions and as quantitative information the number of visits of floral visitors to individual flowers that account for the intensity of the interaction. Although floral visitors are not always pollinators and number of visits does not consider each pollinator species efficiency (\protect\hyperlink{ref-ballantyne2015}{Ballantyne et al., 2015}), the number of visits can provide valuable information of the contribution of floral visitors to pollination (\protect\hyperlink{ref-vazquez2005}{Vázquez et al., 2005}, \protect\hyperlink{ref-vazquez2012}{2012}). Second, we investigated how the main axes of trait variation and individual traits influence plant species' functional roles in the pollination network using a set of complementary interaction network metrics: number of visits, normalized degree and specialization (see below `plant species network roles' section).

@@

We investigated whether different quantitative traits determined different plant species' functional roles in the pollination network using Bayesian modelling and regression trees. For this, we selected simple and complementary species-level network metrics commonly applied in bipartite network studies (\protect\hyperlink{ref-dormann2008}{Dormann et al., 2008}) with a straightforward ecological interpretation relevant to our research goals. The different plant species-level metrics were: (i) sum of visits per plant species; (ii) normalized degree, calculated as the number of links per plant species divided by the total possible number of partners; and (iii) specialization (d') (\protect\hyperlink{ref-bluthgen2006}{Blüthgen et al., 2006}), which measures the deviation of an expected random choice of the available interaction partners and ranges between 0 (maximum generalization) and 1 (maximum specialization). Normalized degree and specialization were calculated with the \emph{specieslevel} function from the R package \emph{bipartite} (\protect\hyperlink{ref-dormann2008}{Dormann et al., 2008}).

@@

The phylogenetically informed principal component analysis captured by the first two and three axes 51.8\% and 70.97\% of the reproductive trait variation, respectively (Fig. 1 and Fig. S6). The first principal component (PC1) represented 26.72\% of the trait variation and indicated a negative correlation between flower number and flower size (`flower number - flower size'). The main contributing traits to PC1 were plant height, flower number, ovule number and flower size (loadings \textgreater{} \textbar0.5\textbar; Table S3) but style length also contributed moderately to PC1 (loading = -0.33). One end of this axis comprised species with high investment in flower number and plant height but small flower size, short style length and low ovule number. For instance, on this end of the spectrum we find the species \emph{Cornus florida} which has approximately a total of 10.000 flowers, an average height of 7.5m, flowers of 3mm wide, a style length of 3.5mm and a total of 2 ovules per flower. The other end of this spectrum comprised species that were short in height and invested in large flowers, long styles, many ovules, but few flowers. For instance, on this side of the axis we find the species \emph{Petunia axillaris} that has approximately 10 flowers per plant, a height of 0.5m, flowers over 50mm wide, styles of 25mm and over 200 ovules per flower. The second principal component (PC2) represented 25.05\% of the trait variation and indicated the variation from low to high autonomous selfing, or in other words, high to low pollinator dependence, respectively. The main driver of trait variation on PC2 was autonomous selfing (loading = 0.85) but the other traits (except ovule number) also made moderate contributions (loadings from 0.27 to 0.4; Table S3). In general terms, species with high pollinator dependence were associated with larger and a higher number of flowers, greater plant height and longer styles. A species example of this side of the spectrum is \emph{Zuccagnia punctata} that is a self incompatible shrub which depends totally on floral visitors, it has approximately 1500 flowers per plant, 3m height and a style length of 20mm. In contrast, species with low pollinator dependence tended to have fewer and smaller flowers, shorter plant height and shorter styles. For instance, on this side of the spectrum we find the species \emph{Veronica peregrina} which is a self compatible herb that is thought to be almost a complete selfer, it has around 20 flowers per plant, a height of 0.2m and a style length of 0.25mm. Further, PC3 explained a considerable amount of trait variability (19.17\%) and the main contributors to this axis were style length (loading = -0.66) and the degree of autonomous selfing (loading = -0.51). The remaining traits, apart from ovule number, were moderately correlated to changes on PC3 (loadings from -0.23 to -0.46; Table S3). In addition, the pPCA with the subset of species that had nectar and pollen quantity data showed that nectar quantity (microlitres of nectar per flower) and pollen grains per flower were positively associated with flower size, style length and ovule number but negatively associated with flower number (PC1, 26.82\%; Fig. S7). This pPCA explained similar variance with the first two principal components (45.52\%) and similar associations of traits despite some variability in the loadings (Table S4).

@@

We found a strong phylogenetic signal (\emph{P} \textless{} 0.01) for most quantitative traits (Table S5). The traits that showed the highest phylogenetic signal were ovule number (\(\lambda\) = 1), pollen grains per flower (\(\lambda\) = 1) and plant height (\(\lambda\) = 0.96), followed by flower length (\(\lambda\) = 0.75), flower width (\(\lambda\) = 0.73), number of flowers per plant (\(\lambda\) = 0.69) and nectar concentration (\(\lambda\) = 0.65). The traits that showed a moderate phylogenetic signal were inflorescence width (\(\lambda\) = 0.57), style length (\(\lambda\) = 0.49) and autonomous selfing (\(\lambda\) = 0.34). Finally, microliters of nectar per flower showed the lowest phylogenetic signal of all traits (\(\lambda\) = 0.14).

@@

\leavevmode\vadjust pre{\hypertarget{ref-serneels2008}{}}%
Serneels, S., \& Verdonck, T. (2008). Principal component analysis for data containing outliers and missing elements. \emph{Computational Statistics \& Data Analysis}, \emph{52}(3), 1712--1727.

@@

\begin{figure}
\centering
\includegraphics{output/figures/unnamed-chunk-4-1.pdf}
\caption{\label{fig:unnamed-chunk-4}\textbf{Figure 3} Fitted posterior estimates of the presence-absence of interaction (a, b and c) and number of visits (d, e and f) of the different floral visitor guilds in relation to the main axes of trait variation (PC1, PC2 and PC3). PC1 represents the flower number - flower size axis, PC2 represents the pollinator dependence axis and PC3, the style length - pollinator dependence axis. For visualization purposes, due to large differences between number of visits of bees and the rest of guilds, the number of visits was log-transformed (Y-axis of lower panel).}
\end{figure}
